### 2025-10-08 YEAR2/PHY293_LectureNotes/main.tex
oct 7

\subsection{Governing Equations of Harmonic Oscillators}
This subsection collects the baseline ODEs for simple, damped, and driven oscillators to set notation used later.

@@

\begin{definition}[Driven Harmonic Oscillator]
    A driven harmonic oscillator is described by the following differential equation, which includes an external driving force $F(t)$:
    \begin{equation}
        m \frac{d^2 x}{dt^2} + b \frac{dx}{dt} + kx = F(t)   
    \end{equation}
\end{definition}
% [Moved: The Wave Equation section is presented later for a smoother flow after oscillators]

\subsection{Simple Harmonic Motion}
We analyze the undamped solution forms, relate constants to initial conditions, and derive period/frequency relations.
\begin{definition}[Simple Harmonic Motion]
    You should have learned Hooke's law and Newton's second law, which give us the equation of motion for a simple harmonic oscillator. The same as equation (\ref{eq:hooke}), which can be rewritten as:
    \begin{equation}
        F = m\ddot{x} = -kx
    \end{equation}
    By setting $\omega^2 = \frac{k}{m}$, a general solution can be written as:
    \begin{equation}
        x(t) = x_0 + A_1 \cos(\omega t) + A_2 \sin(\omega t)
    \end{equation}
    where $A_1$ and $A_2$ are constants determined by the IVP, $\omega$ is the angular frequency, and $\phi$ is the phase constant. $x_0$ is the equilibrium position (often set to 0). The unknown constants can be determined by knowing $x, \dot{x}$ at specific times.
\end{definition}

\begin{definition}[Period, Frequency, and Angular Frequency]
    The period \( T \) is the time it takes for one complete cycle of the motion, given by:
    \begin{equation}
        T = 2\pi \sqrt{\frac{m}{k}}
    \end{equation}
    The frequency \( f \) is the number of cycles per unit time, given by:
    \begin{equation}
        f = \frac{1}{T} = \frac{1}{2\pi} \sqrt{\frac{k}{m}}
    \end{equation}
    The angular frequency \( \omega \) is related to the frequency by:
    \begin{equation}
        \omega = 2\pi f = \sqrt{\frac{k}{m}}
    \end{equation}

@@

    We can find $\omega$ from the given $m$ and $k$:
    $$
    \omega = \sqrt{\frac{k}{m}} = \sqrt{\frac{201\ \mathrm{N\,m^{-1}}}{11.0\ \mathrm{kg}}} \approx 4.28\ \mathrm{rad\,s^{-1}}
    $$
    Therefore, we can solve for $A_2$:
    $$
    A_2 = \frac{v(0)}{\omega} = \frac{-1.33\ \mathrm{m\,s^{-1}}}{4.28\ \mathrm{rad\,s^{-1}}} \approx -0.311\ \mathrm{m}
    $$
    Thus, the equation describing the position \(x(t)\) of the oscillator as a function of time is:
    $$
    x(t) = -0.207 \cos(4.28 t) - 0.311 \sin(4.28 t)
    $$
\end{example}
\begin{theorem}[A Trigonometric Identity]
We can also express the solution in a more compact form using a single cosine function with a phase shift:
\begin{equation}
    x(t) = A \cos(\omega t + \phi)
\end{equation}
where
\begin{subequations}
    \begin{align}
        A &= \sqrt{A_1^2 + A_2^2}, \\
        \phi &= \arctan\!\left(\frac{-A_2}{A_1}\right) 
              = \arctan\!\left(\frac{-v(0)/\omega}{x(0)}\right).
    \end{align}
\end{subequations}
\end{theorem}

@@

\begin{definition}[The Energy of a Simple Harmonic Oscillator]
    The total mechanical energy \(E\) of a simple harmonic oscillator is the sum of its kinetic energy \(K\) and potential energy \(U\). 
    \begin{equation}
        E = K + U
    \end{equation}
    First we consider the change of potential energy from a position $x_i$ to $x_f$, assuming the path is along the spring or the curve $C$ of the oscillator. The force exerted by the spring is given by Hooke's law, \( F = -kx \). The change in potential energy can be simply parametized and calculated as follows:
    \begin{equation}
        \Delta U = \int_C F\cdot ds = -\int_{x_i}^{x_f} F\,dx = \int_{x_i}^{x_f} kx\,dx = \left[\frac{1}{2}kx^2\right]_{x_i}^{x_f} = \frac{1}{2}k(x_f^2 - x_i^2)
    \end{equation}

    Therefore, the potential energy \(U\) at a position \(x\) (taking the reference point at \(x=0\)) is given by:
    \begin{equation}
        U(x) = \frac{1}{2}kx^2
    \end{equation}

    The kinetic energy \(K\) of the oscillator is given by:
    \begin{equation}
        K = \frac{1}{2}m\dot{x}^2
    \end{equation}
    Therefore, the total mechanical energy \(E\) of the simple harmonic oscillator is:
    \begin{equation}
        E = K + U = \frac{1}{2}m\dot{x}^2 + \frac{1}{2}kx^2
    \end{equation}
    The total mechanical energy \(E\) remains constant over time, as energy is conserved in the absence of non-conservative forces (like friction or air resistance).

@@

\subsection{Damped Harmonic Motion}
We solve the damped ODE, classify regimes (underdamped, critical, overdamped), and connect decay rates with parameters.
\begin{definition}[Damped Harmonic Motion]
    For small velocities, the drag force is approximately proportional to the velocity and acts in the opposite direction. This drag force can be modeled as \( F_d = -\gamma \dot{x} \), where \( \gamma \) is the damping coefficient. Including this drag force in the equation of motion for a harmonic oscillator leads to the damped harmonic oscillator equation (\ref{eq:damped_ho}). Which could be rewritten as:
    \begin{equation}
        \ddot{x} + \gamma \dot{x} + \omega_0^2 x = 0
    \end{equation}
    where \( \omega_0 = \sqrt{\frac{k}{m}} \) is the natural angular frequency of the undamped oscillator, and \( \gamma = \frac{b}{m} \) is the damping coefficient per unit mass.

    To skip the math, lets assume a solution of the form \( x(t) = e^{i\omega t} \), substituting into the differential equation gives us a formulation for \( \omega \):
    \begin{align}
        -\omega^2 - i\gamma\omega + \omega_0^2 &= 0 \nonumber \\
        \omega &= -i\frac{\gamma}{2} \pm \sqrt{\omega_0^2 - \frac{\gamma^2}{4}} \label{eq:omega_damped}
    \end{align}
    Also, we can characterize the real and imaginary parts of \( \omega \) as:
    \begin{subequations}
        \begin{equation}
            \omega_r = \mathrm{Re}(\omega) = \pm \sqrt{\omega_0^2 - \frac{\gamma^2}{4}}
        \end{equation}
        \begin{equation}
            \omega_i = \mathrm{Im}(\omega) = -\frac{\gamma}{2}
        \end{equation}
    \end{subequations}
    The general solution for the damped harmonic oscillator can be written as:
    \begin{equation}
        x(t) = \exp(\omega_i t) \exp(-i \omega_r t) = \exp\left(-\frac{\gamma}{2}t\right) \exp(\mp i \sqrt{\omega_0^2 - \frac{\gamma^2}{4}} t)
    \end{equation}
    \begin{subequations}
    \begin{itemize}
        \item \textbf{No Damping} (\( \gamma = 0 \)): The system behaves like a simple harmonic oscillator with angular frequency \( \omega_0 \). Given by:
        \begin{equation}
            z = \exp(-i\omega_0 t)
        \end{equation}
        \item \textbf{Underdamping} (\( 0 < \gamma < 2\omega_0 \)): The system oscillates with a gradually decreasing amplitude. The angular frequency of oscillation is given by \( \omega_r = \sqrt{\omega_0^2 - \frac{\gamma^2}{4}} \). Given by:
        \begin{equation}
            z = \exp\left(-\frac{\gamma}{2}t\right) \exp(-i \omega_r t)
        \end{equation}
        The trigonometric form of the solution is:
        \begin{equation}
            x(t) = A_0 \exp\left(-\frac{\gamma}{2}t\right) \cos(\omega_r t + \phi)
        \end{equation}
        where \( A_0 \) and \( \phi \) are constants determined by the initial conditions
        From this, we can derive the following cases:
        \item \textbf{Critical Damping} (\( \gamma = 2\omega_0 \)): The system returns to equilibrium as quickly as possible without oscillating. Consider:
        $$
            x(t) = e^{-\frac{\gamma}{2}t}f(t)
        $$
        Inserting into the differential equation, we get:
        $$
            \ddot{f} + \left(\omega_0^2 - \frac{\gamma^2}{4}\right)f = 0
        $$
        Since \( \gamma = 2\omega_0 \), we have \( \omega_0^2 - \frac{\gamma^2}{4} = 0 \), leading to:
        $$
            \ddot{f} = 0 \implies f(t) = A_1 t + A_2
        $$
        Therefore, the general solution for the critically damped case is:
        \begin{equation}
            x(t) = (A_1 t + A_2) \exp\left(-\frac{\gamma}{2}t\right)
        \end{equation}
        where \( A_1 \) and \( A_2 \) are constants determined by
        \item \textbf{Overdamping} (\( \gamma > 2\omega_0 \)): The system returns to equilibrium without oscillating, but more slowly than in the critically damped case. The solution is given by:
        \begin{equation}
            z = \exp\left(-\frac{\gamma}{2}t\right) \exp\left(\sqrt{\frac{\gamma^2}{4} - \omega_0^2} t\right)
        \end{equation}
        So the general solution is (the solution is via a substitution of $x(t) = e^{-\gamma t/2} f(t)$ into the differential equation, which resolves the ODE to a simple form):
        \begin{equation}
            x(t) = A_1 \exp\left[\left(-\frac{\gamma}{2} + \sqrt{\frac{\gamma^2}{4} - \omega_0^2}\right) t\right] + A_2 \exp\left[\left(-\frac{\gamma}{2} - \sqrt{\frac{\gamma^2}{4} - \omega_0^2}\right) t\right]
        \end{equation}
        where \( A_1 \) and \( A_2 \) are constants determined by the initial conditions.
    \end{itemize}
    \end{subequations}
\end{definition}

\subsection{Energy and Quality Factor}
We study how energy decays under light damping, define the time constant and quality factor Q, and relate them to response.
\begin{definition}[Energy of a Very Light Damping]
    Consider a very lightly damped harmonic oscillator, where \( \gamma \ll \omega_0 \). In this case, the angular frequency of oscillation \( \omega_r \) can be approximated as:
    $$
    \omega_r \approx \omega_0 \left(1 - \frac{\gamma^2}{8\omega_0^2}\right) \approx \omega_0
    $$
    So the motion of the lightly damped oscillator can be approximated as:
    $$
    x(t) \approx A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi)
    $$
    Then, we can calculate the velocity of the oscillator:
    \begin{align*}
        \dot{x}(t) &= -\frac{\gamma}{2} A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi) - A_0 \omega_0 e^{-\frac{\gamma}{2}t} \sin(\omega_0 t + \phi) \\
        &= A_0 \omega_0 e^{-\frac{\gamma}{2}t} \left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)
    \end{align*}
    The total mechanical energy \( E(t) \) of the lightly damped oscillator is given by:
    \begin{align*}
        E(t) &= \frac{1}{2}m\dot{x}^2 + \frac{1}{2}kx^2 \\
             &= \frac{1}{2}m \left[A_0 \omega_0 e^{-\frac{\gamma}{2}t} \left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)\right]^2 + \frac{1}{2}k \left[A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi)\right]^2 \\
             &= \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[\left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)^2 + \cos^2(\omega_0 t + \phi)\right] \\
             &= \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[\sin^2(\omega_0 t + \phi) + \cos^2(\omega_0 t + \phi) + \frac{\gamma^2}{4\omega_0^2} \cos^2(\omega_0 t + \phi) + \frac{\gamma}{\omega_0} \sin(\omega_0 t + \phi) \cos(\omega_0 t + \phi)\right] \\
             &\approx \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[1 + \frac{\gamma^2}{4\omega_0^2} \cos^2(\omega_0 t + \phi)\right] \quad (\text{neglecting the small term } \frac{\gamma}{\omega_0} \sin(\omega_0 t + \phi) \cos(\omega_0 t + \phi)) \\
             &\approx \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \quad (\text{since } \frac{\gamma^2}{4\omega_0^2} \text{ is very small}) \\
             &= E_0 e^{-\gamma t} \quad \text{where } E_0 = \frac{1}{2}m A_0^2 \omega_0^2 \text{ is the initial energy at } t=0
    \end{align*}
    We can also define the time constant \( \tau \) as the time it takes for the energy to decrease to \( \frac{1}{e} \) of its initial value:
    \begin{equation}
        \tau = \frac{1}{\gamma}
    \end{equation}
    So we have, for very light damping:
    \begin{equation}
        E(t) = E_0 e^{-\gamma t} = E_0 e^{-\frac{t}{\tau}}
    \end{equation}
\end{definition}

\begin{definition}[Rate of Energy Loss]
    Taking the time derivative of the total mechanical energy \( E(t) \):
    \begin{align*}
        \frac{dE}{dt} &= \frac{d}{dt} \left(\frac{1}{2}m\dot{x}^2 + \frac{1}{2}kx^2 \right) \\
                      &= (ma + kx)\dot{x} \\
    \end{align*}
    For a undamped harmonic oscillator, \( ma + kx = 0 \), so \( \frac{dE}{dt} = 0 \), indicating that the total mechanical energy is conserved and obeys Hooke's law completely. However, for a damped harmonic oscillator, \( ma + kx = -b\dot{x} \), leading to:
    \begin{equation}
        \frac{dE}{dt} = -b\dot{x}^2
    \end{equation}
\end{definition}

\begin{definition}[Quality Factor (Q-Factor)]
    The quality factor \( Q \) is a dimensionless parameter that characterizes the damping of a harmonic oscillator. It is defined as:
    \begin{subequations}
    \begin{equation}
        Q = \frac{\omega}{\gamma} = \omega \tau
    \end{equation}
    And for very light damping, we can approximate \( \omega \approx \omega_0 \), leading to:
    \begin{equation}
        Q \approx \frac{\omega_0}{\gamma} = \omega_0 \tau
    \end{equation}
    \end{subequations}
    This allows us to rewrite the equation of a damped harmonic oscillator as:
    \begin{equation}
        \ddot{x} + \frac{\omega_0}{Q} \dot{x} + \omega_0^2 x = 0
    \end{equation}
    and:
    \begin{equation}
        \omega = \omega_0 \sqrt{1 - \frac{1}{4Q^2}}
    \end{equation}

@@

\subsection{Undamped Forced Oscillations}
We examine steady-state response to sinusoidal driving without losses and identify resonance behavior.
\begin{definition}[Undamped Forced Oscillations]
    Consider a driver force acting on the mass-spring system; the ODE becomes:
    \begin{equation}
        m\ddot{x} + kx = \eta
    \end{equation}
    where \( \eta = F_0 \cos(\omega t) \) is the driving force with amplitude \( F_0 \) and angular frequency \( \omega \). The steady-state particular solution is:
    \begin{equation}
        x(t) = A(\omega)\cos(\omega t - \delta) 
    \end{equation}
    To derive $A(\omega)$ and $\delta$, use the equation of motion:
    \begin{align*}
        A(\omega)(-\omega^2 + \omega_0^2) \cos(\delta) = \omega_0^2 a
    \end{align*}
    where \( a = \frac{F_0}{k} \) is the static displacement of the mass when the driving force is constant. We also have:
    \begin{align*}
        A(\omega)(-\omega^2 + \omega_0^2) \sin(\delta) = 0
    \end{align*}
    Now, consider the case when $\delta = 0$, we have:
    \begin{equation}
        A(\omega) = \frac{\omega_0^2 a}{\omega_0^2 - \omega^2}
    \end{equation}
    and the case where \( \delta = \pi \), we have:
    \begin{equation}
        A(\omega) = -\frac{\omega_0^2 a}{\omega_0^2 - \omega^2}
    \end{equation}
    Observe, when $\omega \approx \omega_0$, the amplitude \( A(\omega) \) becomes very large, indicating resonance. At resonance, the system oscillates with maximum amplitude, which can lead to significant energy transfer from the driving force to the oscillator.
\end{definition}

>>>>

\subsection{Governing Equations of Harmonic Oscillators}

\begin{keybox}
\textbf{Key Equations:}
\begin{itemize}
    \item Simple Harmonic Oscillator: $m\ddot{x} + kx = 0$ or $\ddot{x} + \omega_0^2 x = 0$ where $\omega_0 = \sqrt{k/m}$
    \item Damped: $m\ddot{x} + b\dot{x} + kx = 0$ or $\ddot{x} + \gamma\dot{x} + \omega_0^2 x = 0$ where $\gamma = b/m$
    \item Driven: $m\ddot{x} + b\dot{x} + kx = F(t)$
\end{itemize}
\textbf{Tip:} Remember to check the damping parameter convention ($\gamma$ vs $2\beta$) when comparing formulas!
\end{keybox}

This subsection collects the baseline ODEs for simple, damped, and driven oscillators to set notation used later.

@@

\begin{definition}[Driven Harmonic Oscillator]
    A driven harmonic oscillator is described by the following differential equation, which includes an external driving force $F(t)$:
    $$
        m \frac{d^2 x}{dt^2} + b \frac{dx}{dt} + kx = F(t)   
    $$
\end{definition}
% [Moved: The Wave Equation section is presented later for a smoother flow after oscillators]

\subsection{Simple Harmonic Motion}

\begin{keybox}
\textbf{Key Equations:}
\begin{itemize}
    \item General solution: $x(t) = A\cos(\omega t + \phi)$ where $\omega = \sqrt{k/m}$
    \item Period: $T = 2\pi/\omega = 2\pi\sqrt{m/k}$
    \item Frequency: $f = 1/T = \omega/(2\pi)$
    \item Total Energy: $E = \frac{1}{2}kA^2$ (constant for undamped oscillator)
\end{itemize}
\textbf{Common Pitfall:} Don't forget to match initial conditions properly when solving for amplitude and phase!
\end{keybox}

We analyze the undamped solution forms, relate constants to initial conditions, and derive period/frequency relations.
\begin{definition}[Simple Harmonic Motion]
    You should have learned Hooke's law and Newton's second law, which give us the equation of motion for a simple harmonic oscillator. The same as equation (\ref{eq:hooke}), which can be rewritten as:
    $$
        F = m\ddot{x} = -kx
    $$
    By setting $\omega^2 = \frac{k}{m}$, a general solution can be written as:
    $$
        x(t) = x_0 + A_1 \cos(\omega t) + A_2 \sin(\omega t)
    $$
    where $A_1$ and $A_2$ are constants determined by the IVP, $\omega$ is the angular frequency, and $\phi$ is the phase constant. $x_0$ is the equilibrium position (often set to 0). The unknown constants can be determined by knowing $x, \dot{x}$ at specific times.
\end{definition}

\begin{definition}[Period, Frequency, and Angular Frequency]
    The period \( T \) is the time it takes for one complete cycle of the motion, given by:
    $$
        T = 2\pi \sqrt{\frac{m}{k}}
    $$
    The frequency \( f \) is the number of cycles per unit time, given by:
    $$
        f = \frac{1}{T} = \frac{1}{2\pi} \sqrt{\frac{k}{m}}
    $$
    The angular frequency \( \omega \) is related to the frequency by:
    $$
        \omega = 2\pi f = \sqrt{\frac{k}{m}}
    $$

@@

    We can find $\omega$ from the given $m$ and $k$:
    $$
    \omega = \sqrt{\frac{k}{m}} = \sqrt{\frac{201\ \mathrm{N\,m^{-1}}}{11.0\ \mathrm{kg}}} \approx 4.28\ \mathrm{rad\,s^{-1}}
    $$
    Therefore, we can solve for $A_2$:
    $$
    A_2 = \frac{v(0)}{\omega} = \frac{-1.33\ \mathrm{m\,s^{-1}}}{4.28\ \mathrm{rad\,s^{-1}}} \approx -0.311\ \mathrm{m}
    $$
    Thus, the equation describing the position \(x(t)\) of the oscillator as a function of time is:
    $$
    x(t) = -0.207 \cos(4.28 t) - 0.311 \sin(4.28 t)
    $$
\end{example}
\begin{theorem}[A Trigonometric Identity]
We can also express the solution in a more compact form using a single cosine function with a phase shift:
$$
    x(t) = A \cos(\omega t + \phi)
$$
where
$$
        A = \sqrt{A_1^2 + A_2^2}, \qquad
        \phi = \arctan\!\left(\frac{-A_2}{A_1}\right) 
              = \arctan\!\left(\frac{-v(0)/\omega}{x(0)}\right)
$$
\end{theorem}

@@

\begin{definition}[The Energy of a Simple Harmonic Oscillator]
    The total mechanical energy \(E\) of a simple harmonic oscillator is the sum of its kinetic energy \(K\) and potential energy \(U\). 
    $$
        E = K + U
    $$
    First we consider the change of potential energy from a position $x_i$ to $x_f$, assuming the path is along the spring or the curve $C$ of the oscillator. The force exerted by the spring is given by Hooke's law, \( F = -kx \). The change in potential energy can be simply parametized and calculated as follows:
    $$
        \Delta U = \int_C F\cdot ds = -\int_{x_i}^{x_f} F\,dx = \int_{x_i}^{x_f} kx\,dx = \left[\frac{1}{2}kx^2\right]_{x_i}^{x_f} = \frac{1}{2}k(x_f^2 - x_i^2)
    $$

    Therefore, the potential energy \(U\) at a position \(x\) (taking the reference point at \(x=0\)) is given by:
    $$
        U(x) = \frac{1}{2}kx^2
    $$

    The kinetic energy \(K\) of the oscillator is given by:
    $$
        K = \frac{1}{2}m\dot{x}^2
    $$
    Therefore, the total mechanical energy \(E\) of the simple harmonic oscillator is:
    \begin{equation}\label{eq:energy_sho}
        E = K + U = \frac{1}{2}m\dot{x}^2 + \frac{1}{2}kx^2
    \end{equation}
    The total mechanical energy \(E\) remains constant over time, as energy is conserved in the absence of non-conservative forces (like friction or air resistance).

@@

\subsection{Damped Harmonic Motion}

\begin{keybox}
\textbf{Key Equations:}
\begin{itemize}
    \item Damped ODE: $\ddot{x} + \gamma\dot{x} + \omega_0^2 x = 0$ where $\gamma = b/m$
    \item Underdamped ($\gamma < 2\omega_0$): $x(t) = A_0 e^{-\gamma t/2}\cos(\omega_r t + \phi)$ where $\omega_r = \sqrt{\omega_0^2 - \gamma^2/4}$
    \item Critically damped ($\gamma = 2\omega_0$): $x(t) = (A_1 t + A_2)e^{-\gamma t/2}$
    \item Overdamped ($\gamma > 2\omega_0$): Exponential decay without oscillation
\end{itemize}
\textbf{Concept:} Critical damping returns to equilibrium fastest without oscillating!
\end{keybox}

We solve the damped ODE, classify regimes (underdamped, critical, overdamped), and connect decay rates with parameters.
\begin{definition}[Damped Harmonic Motion]
    For small velocities, the drag force is approximately proportional to the velocity and acts in the opposite direction. This drag force can be modeled as \( F_d = -\gamma \dot{x} \), where \( \gamma \) is the damping coefficient. Including this drag force in the equation of motion for a harmonic oscillator leads to the damped harmonic oscillator equation (\ref{eq:damped_ho}). Which could be rewritten as:
    $$
        \ddot{x} + \gamma \dot{x} + \omega_0^2 x = 0
    $$
    where \( \omega_0 = \sqrt{\frac{k}{m}} \) is the natural angular frequency of the undamped oscillator, and \( \gamma = \frac{b}{m} \) is the damping coefficient per unit mass.

    To skip the math, lets assume a solution of the form \( x(t) = e^{i\omega t} \), substituting into the differential equation gives us a formulation for \( \omega \):
    \begin{equation}\label{eq:omega_damped}
        \omega = -i\frac{\gamma}{2} \pm \sqrt{\omega_0^2 - \frac{\gamma^2}{4}}
    \end{equation}
    Also, we can characterize the real and imaginary parts of \( \omega \) as:
    $$
        \omega_r = \mathrm{Re}(\omega) = \pm \sqrt{\omega_0^2 - \frac{\gamma^2}{4}}, \qquad
        \omega_i = \mathrm{Im}(\omega) = -\frac{\gamma}{2}
    $$
    The general solution for the damped harmonic oscillator can be written as:
    $$
        x(t) = \exp(\omega_i t) \exp(-i \omega_r t) = \exp\left(-\frac{\gamma}{2}t\right) \exp(\mp i \sqrt{\omega_0^2 - \frac{\gamma^2}{4}} t)
    $$
    \begin{subequations}
    \begin{itemize}
        \item \textbf{No Damping} (\( \gamma = 0 \)): The system behaves like a simple harmonic oscillator with angular frequency \( \omega_0 \). Given by:
        $$
            z = \exp(-i\omega_0 t)
        $$
        \item \textbf{Underdamping} (\( 0 < \gamma < 2\omega_0 \)): The system oscillates with a gradually decreasing amplitude. The angular frequency of oscillation is given by \( \omega_r = \sqrt{\omega_0^2 - \frac{\gamma^2}{4}} \). Given by:
        $$
            z = \exp\left(-\frac{\gamma}{2}t\right) \exp(-i \omega_r t)
        $$
        The trigonometric form of the solution is:
        $$
            x(t) = A_0 \exp\left(-\frac{\gamma}{2}t\right) \cos(\omega_r t + \phi)
        $$
        where \( A_0 \) and \( \phi \) are constants determined by the initial conditions
        From this, we can derive the following cases:
        \item \textbf{Critical Damping} (\( \gamma = 2\omega_0 \)): The system returns to equilibrium as quickly as possible without oscillating. Consider:
        $$
            x(t) = e^{-\frac{\gamma}{2}t}f(t)
        $$
        Inserting into the differential equation, we get:
        $$
            \ddot{f} + \left(\omega_0^2 - \frac{\gamma^2}{4}\right)f = 0
        $$
        Since \( \gamma = 2\omega_0 \), we have \( \omega_0^2 - \frac{\gamma^2}{4} = 0 \), leading to:
        $$
            \ddot{f} = 0 \implies f(t) = A_1 t + A_2
        $$
        Therefore, the general solution for the critically damped case is:
        $$
            x(t) = (A_1 t + A_2) \exp\left(-\frac{\gamma}{2}t\right)
        $$
        where \( A_1 \) and \( A_2 \) are constants determined by
        \item \textbf{Overdamping} (\( \gamma > 2\omega_0 \)): The system returns to equilibrium without oscillating, but more slowly than in the critically damped case. The solution is given by:
        $$
            z = \exp\left(-\frac{\gamma}{2}t\right) \exp\left(\sqrt{\frac{\gamma^2}{4} - \omega_0^2} t\right)
        $$
        So the general solution is (the solution is via a substitution of $x(t) = e^{-\gamma t/2} f(t)$ into the differential equation, which resolves the ODE to a simple form):
        $$
            x(t) = A_1 \exp\left[\left(-\frac{\gamma}{2} + \sqrt{\frac{\gamma^2}{4} - \omega_0^2}\right) t\right] + A_2 \exp\left[\left(-\frac{\gamma}{2} - \sqrt{\frac{\gamma^2}{4} - \omega_0^2}\right) t\right]
        $$
        where \( A_1 \) and \( A_2 \) are constants determined by the initial conditions.
    \end{itemize}
    \end{subequations}
\end{definition}

\subsection{Energy and Quality Factor}

\begin{keybox}
\textbf{Key Equations:}
\begin{itemize}
    \item Energy decay: $E(t) = E_0 e^{-\gamma t}$ for light damping
    \item Time constant: $\tau = 1/\gamma$ (time for energy to drop to $1/e$)
    \item Quality factor: $Q = \omega_0/\gamma = \omega_0\tau$ (higher $Q$ = less damping)
    \item Energy loss per cycle: $\frac{\Delta E}{E} \approx -\frac{2\pi}{Q}$
\end{itemize}
\textbf{Concept:} $Q$ measures how many oscillations occur before energy decays significantly!
\end{keybox}

We study how energy decays under light damping, define the time constant and quality factor Q, and relate them to response.
\begin{definition}[Energy of a Very Light Damping]
    Consider a very lightly damped harmonic oscillator, where \( \gamma \ll \omega_0 \). In this case, the angular frequency of oscillation \( \omega_r \) can be approximated as:
    $$
    \omega_r \approx \omega_0 \left(1 - \frac{\gamma^2}{8\omega_0^2}\right) \approx \omega_0
    $$
    So the motion of the lightly damped oscillator can be approximated as:
    $$
    x(t) \approx A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi)
    $$
    Then, we can calculate the velocity of the oscillator:
    \begin{align*}
        \dot{x}(t) &= -\frac{\gamma}{2} A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi) - A_0 \omega_0 e^{-\frac{\gamma}{2}t} \sin(\omega_0 t + \phi) \\
        &= A_0 \omega_0 e^{-\frac{\gamma}{2}t} \left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)
    \end{align*}
    The total mechanical energy \( E(t) \) of the lightly damped oscillator is given by:
    \begin{align*}
        E(t) &= \frac{1}{2}m\dot{x}^2 + \frac{1}{2}kx^2 \\
             &= \frac{1}{2}m \left[A_0 \omega_0 e^{-\frac{\gamma}{2}t} \left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)\right]^2 + \frac{1}{2}k \left[A_0 e^{-\frac{\gamma}{2}t} \cos(\omega_0 t + \phi)\right]^2 \\
             &= \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[\left(-\frac{\gamma}{2\omega_0} \cos(\omega_0 t + \phi) - \sin(\omega_0 t + \phi)\right)^2 + \cos^2(\omega_0 t + \phi)\right] \\
             &= \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[\sin^2(\omega_0 t + \phi) + \cos^2(\omega_0 t + \phi) + \frac{\gamma^2}{4\omega_0^2} \cos^2(\omega_0 t + \phi) + \frac{\gamma}{\omega_0} \sin(\omega_0 t + \phi) \cos(\omega_0 t + \phi)\right] \\
             &\approx \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \left[1 + \frac{\gamma^2}{4\omega_0^2} \cos^2(\omega_0 t + \phi)\right] \quad (\text{neglecting the small term } \frac{\gamma}{\omega_0} \sin(\omega_0 t + \phi) \cos(\omega_0 t + \phi)) \\
             &\approx \frac{1}{2}m A_0^2 \omega_0^2 e^{-\gamma t} \quad (\text{since } \frac{\gamma^2}{4\omega_0^2} \text{ is very small}) \\
             &= E_0 e^{-\gamma t} \quad \text{where } E_0 = \frac{1}{2}m A_0^2 \omega_0^2 \text{ is the initial energy at } t=0
    \end{align*}
    We can also define the time constant \( \tau \) as the time it takes for the energy to decrease to \( \frac{1}{e} \) of its initial value:
    $$
        \tau = \frac{1}{\gamma}
    $$
    So we have, for very light damping:
    \begin{equation}\label{eq:energy_decay}
        E(t) = E_0 e^{-\gamma t} = E_0 e^{-\frac{t}{\tau}}
    \end{equation}
\end{definition}

\begin{definition}[Rate of Energy Loss]
    Taking the time derivative of the total mechanical energy \( E(t) \):
    \begin{align*}
        \frac{dE}{dt} &= \frac{d}{dt} \left(\frac{1}{2}m\dot{x}^2 + \frac{1}{2}kx^2 \right) \\
                      &= (ma + kx)\dot{x} \\
    \end{align*}
    For a undamped harmonic oscillator, \( ma + kx = 0 \), so \( \frac{dE}{dt} = 0 \), indicating that the total mechanical energy is conserved and obeys Hooke's law completely. However, for a damped harmonic oscillator, \( ma + kx = -b\dot{x} \), leading to:
    $$
        \frac{dE}{dt} = -b\dot{x}^2
    $$
\end{definition}

\begin{definition}[Quality Factor (Q-Factor)]
    The quality factor \( Q \) is a dimensionless parameter that characterizes the damping of a harmonic oscillator. It is defined as:
    \begin{equation}\label{eq:Q_factor}
        Q = \frac{\omega}{\gamma} = \omega \tau
    \end{equation}
    And for very light damping, we can approximate \( \omega \approx \omega_0 \), leading to:
    $$
        Q \approx \frac{\omega_0}{\gamma} = \omega_0 \tau
    $$
    This allows us to rewrite the equation of a damped harmonic oscillator as:
    $$
        \ddot{x} + \frac{\omega_0}{Q} \dot{x} + \omega_0^2 x = 0
    $$
    and:
    $$
        \omega = \omega_0 \sqrt{1 - \frac{1}{4Q^2}}
    $$

@@

\subsection{Undamped Forced Oscillations}

\begin{keybox}
\textbf{Key Equations:}
\begin{itemize}
    \item Driven ODE: $m\ddot{x} + kx = F_0\cos(\omega t)$
    \item Steady-state solution: $x(t) = A(\omega)\cos(\omega t - \delta)$
    \item Amplitude: $A(\omega) = \frac{\omega_0^2 a}{|\omega_0^2 - \omega^2|}$ where $a = F_0/k$
\end{itemize}
\textbf{Common Pitfall:} At resonance $(\omega \approx \omega_0)$, amplitude becomes infinite without damping!
\end{keybox}

We examine steady-state response to sinusoidal driving without losses and identify resonance behavior.
\begin{definition}[Undamped Forced Oscillations]
    Consider a driver force acting on the mass-spring system; the ODE becomes:
    $$
        m\ddot{x} + kx = \eta
    $$
    where \( \eta = F_0 \cos(\omega t) \) is the driving force with amplitude \( F_0 \) and angular frequency \( \omega \). The steady-state particular solution is:
    $$
        x(t) = A(\omega)\cos(\omega t - \delta) 
    $$
    To derive $A(\omega)$ and $\delta$, use the equation of motion:
    \begin{align*}
        A(\omega)(-\omega^2 + \omega_0^2) \cos(\delta) = \omega_0^2 a
    \end{align*}
    where \( a = \frac{F_0}{k} \) is the static displacement of the mass when the driving force is constant. We also have:
    \begin{align*}
        A(\omega)(-\omega^2 + \omega_0^2) \sin(\delta) = 0
    \end{align*}
    Now, consider the case when $\delta = 0$, we have:
    $$
        A(\omega) = \frac{\omega_0^2 a}{\omega_0^2 - \omega^2}
    $$
    and the case where \( \delta = \pi \), we have:
    $$
        A(\omega) = -\frac{\omega_0^2 a}{\omega_0^2 - \omega^2}
    $$
    Observe, when $\omega \approx \omega_0$, the amplitude \( A(\omega) \) becomes very large, indicating resonance. At resonance, the system oscillates with maximum amplitude, which can lead to significant energy transfer from the driving force to the oscillator.
\end{definition}

\subsection{Damped Forced Oscillations}
